---
output: rmarkdown::github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(
  fig.path = "./man/figures/",
  message = FALSE,
  warning = FALSE,
  collapse = TRUE,
  comment = ""
)

library(devtools)
devtools::load_all(".")
```

# hypeR.GEM

![Gitter](https://img.shields.io/gitter/room/montilab/hypeR-GEM)
![GitHub issues](https://img.shields.io/github/issues/montilab/hypeR-GEM)
![GitHub last commit](https://img.shields.io/github/last-commit/montilab/hypeR-GEM)

`hypeR.GEM` maps metabolite signatures to enzyme-coding genes using genome-scale
metabolic models (GEMs), then performs gene set enrichment analysis in the
mapped gene space. The package is designed for metabolomics studies where the
goal is to infer reaction-based links between metabolites and enzyme-coding
genes, enabling the mapping of metabolite signatures to gene signatures and
their subsequent annotation via gene set enrichment analysis.

## What the package does

- Maps RefMet-annotated metabolites to enzyme-coding genes through GEMs
- Supports non-directional and directional reaction-based mapping
- Quantifies gene-level specificity to reduce the impact of promiscuous genes
- Performs unweighted or weighted hypergeometric enrichment analysis
- Provides plotting and export utilities for downstream interpretation

## Installation

Install the development version from GitHub:

```r
library(devtools)
devtools::install_github("montilab/hypeR-GEM")
```

Then load the package:

```r
library(hypeR.GEM)
```

## Quick start

This example uses metabolite signatures from human urine samples in COVID-19.
By the end of this example, you will map metabolites to enzyme-coding genes and
identify enriched biological pathways from the mapped gene signatures.

### 1. Load example data

`COVID_urine` is a named list with two signatures:

- `up`: up-regulated metabolites
- `dn`: down-regulated metabolites

Each signature is a data frame containing a RefMet annotation column named
`refmet_name`.

```{r}
data(COVID_urine)
str(COVID_urine, max.level = 1)
```

### 2. Check the expected input format

`signature2gene()` expects a named list of data frames. Each data frame should
contain a column matching `reference_key` such as `refmet_name`.

```{r}
head(COVID_urine$up)
```

### 3. Map metabolites to genes with `signature2gene()`

The main function in the package is `signature2gene()`.

Key arguments:

- `signatures`: named list of metabolic signatures
- `species`: built-in GEM species (`"Human"`, `"Mouse"`, `"Rat"`,
  `"Zebrafish"`, `"Worm"`, or `"Other"`)
- `directional`: if `TRUE`, only reactions producing the metabolite are used;
  if `FALSE`, both reactant and product relationships are used
- `promiscuous_threshold`: filters metabolites associated with many genes
- `reference_key`: metabolite annotation column used for matching to the GEM
- `background`: background size used in the gene-specific hypergeometric test;
  by default, this is set to the total number of metabolites represented in the
  GEM

```{r}
## Non-directional mapping
hypeR_GEM_obj <- hypeR.GEM::signature2gene(signatures = COVID_urine,
                                           directional = FALSE,
                                           promiscuous_threshold = 500,
                                           reference_key = 'refmet_name',
                                           background = NULL)


## Directional mapping
hypeR_GEM_obj_di <- hypeR.GEM::signature2gene(signatures = COVID_urine,
                                              directional = TRUE,
                                              promiscuous_threshold = 500,
                                              reference_key = 'refmet_name',
                                              background = NULL)
```

The returned object contains:

- `promiscuous_met`: metabolites filtered as overly connected
- `mapped_metabolite_signatures`: metabolites matched in the GEM
- `gene_tables`: ranked enzyme-coding genes for each mapeed enzyme-coding signature

```{r}
names(hypeR_GEM_obj)
head(hypeR_GEM_obj$mapped_metabolite_signatures$up)
head(hypeR_GEM_obj$gene_tables$up)
```

### 4. Run gene set enrichment analysis

Here we use the REACTOME gene sets as an example. The background used in this
analysis is set to all enzyme-coding genes in the Human GEM model (for more
details, see [here](https://www.pnas.org/doi/abs/10.1073/pnas.2102344118)).

```{r}
data(reactome)
```


#### 4.1 Unweighted (standard) hypergeometric test
```{r}
## Enrichment analysis from non-directional mapping
enrichment_obj <- hypeR.GEM::enrichment(hypeR_GEM_obj,
                                        genesets = reactome,
                                        genesets_name = "REACTOME",
                                        method='unweighted',
                                        min_metabolite = 2,
                                        background=3068)

## Enrichment analysis from directional mapping
enrichment_obj_di <- hypeR.GEM::enrichment(hypeR_GEM_obj_di,
                                           genesets = reactome,
                                           genesets_name = "REACTOME",
                                           method='unweighted',
                                           min_metabolite = 2,
                                           background=3068)
```

The enrichment result is a list with one element per input signature. Each
element contains:

- `info`: metadata about the test
- `data`: pathway-level enrichment results

```{r}
## Taking the enrichment object from non-directional mapping as an example
names(enrichment_obj)
head(enrichment_obj$up$data)
```

#### 4.2 Weighted hypergeometric test

`hypeR.GEM` also supports weighted enrichment, in which gene-specificity scores
are used to weight genes before testing for pathway enrichment. For a more
detailed explanation of the weighting scheme and workflow, see
`vignette("Workflow", package = "hypeR.GEM")`.

```r
enrichment_obj_wt <- hypeR.GEM::enrichment(hypeR_GEM_obj,
                                           genesets = reactome,
                                           genesets_name = "REACTOME",
                                           method='weighted',
                                           weighted_by = 'fdr',
                                           sigmoid_transformation = TRUE,
                                           min_metabolite = 2,
                                           background=3068)
```

### 5. Visualize the results {.tabset}

```{r}
max_fdr <- 0.05
```


#### 5.1 Non-directional
```{r}
hypeR.GEM::enrichment_plot(enrichment_obj,
                           top=40,
                           abrv=50,
                           size_by="significance",
                           fdr_cutoff= max_fdr,
                           val='fdr') +
  ggplot2::ggtitle(paste("FDR ≤", max_fdr))

```


#### 5.2 Directional
```{r}
hypeR.GEM::enrichment_plot(enrichment_obj_di,
                           top=40,
                           abrv=50,
                           size_by="significance",
                           fdr_cutoff= max_fdr,
                           val='fdr')+
  ggplot2::ggtitle(paste("FDR ≤", max_fdr))
```


## Learn more

For a fuller tutorial, methodological background, and a more detailed workflow,
see the package vignette:

- `vignette("Workflow", package = "hypeR.GEM")`

Related publication:

- [Huang et al.](https://www.biorxiv.org/content/10.64898/2025.12.08.692998v1)
